Draft-assinment1

### 2017-02-21 Draft-xuezhou-sjb.Rmd
added findings

The data is collected by United States Department of Education, which corresponds to all undergraduate degree-granting institutions of higher education from 1996 through 2016. The dataset is pretty rich and covers a wide range of topics. It has 1743 associated columns. Each corresponds to one named data variable. All data variables are divided into 10 categories ranging from ‘Root’ category to ‘Earnings’ category. The data type contains integer, string and float. Data elements with an ‘IPEDS’ value in the ‘Source’ field are derived directly from the Integrated Postsecondary Education Data System (IPEDS). For further detailed description of the dataset, you can find it in the CollegeScorecardDataDictionary.xlsx excel file.

@@

#need to be commented out
mean(is.na(data_new$NPT4_PUB))
```

@@

for (i in 1:length(data_new$NPT4_PRIV)) {
   if (is.na(data_new$NPT4_PRIV[i])) {
       data_new$NPT4_PRIV[i] = data_new$NPT4_PUB[i]
   }
}
mean(is.na(data_new$NPT4_PRIV))
data_new = data_new[,-10] 
colnames(data_new)[10] <- "NPT4_COMBINE"

@@

In order to choose the variables that have enough valid data related to the topic, we filtered and considered variables with percent missing value less than 25%. The new data set was named as ???data_clean??? for further analysis. 
```{r, message=FALSE,warning=FALSE}

@@

For numeric variabls, we calculate the basic summary statistics such as median, min, max, mode and quantile. We also use visualizations such as hisgroms and boxplots to understand the distribution of the values. For categorical variables, we use frequency table to understand the frequcies of each categories.Box chart is used as a visualization tool.

The table below lists the variables that will be used for the further analysis: 12 of them are continuous variables while 3 of them are categorical variables. Our data set now contains 120 rows and 15 columns. 120 rows consist of 20 years??? of data for each of the chosen six countries. The 15 columns consist of country name, country code, year, and the twelve indicators.

| Variable Name    | Data Type |  Unit  |  Description                                                          |    Category in Documentation       |
| --------------   | ----------|--------|-----------------------------------------------------------------------| --------------------------------|
| NPT4_COMBINE    | Numeric   | $         |average net price for TITLE IV institutions of both public and private.|  Cost                       |
| DEP_STAT_PCT_IND | Numeric  | $         |Percentage of students who are financially independent                |     Cost                        |
| PAR_ED_PCT_1STGEN| Numeric  | %         |Percentage first-generation students                                  |      Sudent Body                |
| MD_FAMINC        | Numeric  | $         |Median family income in real 2015 dollars                             |       Earning                   |
| PCTPELL          | Numeric  | %         |Percentage of undergraduates who receive a Pell Grant                 |        Financial Aid            |
| PCTFLOAN         | Numeric  | %         |Percent of all federal undergraduate students receiving a federal student loan      |  Financial Aid    |
| GRAD_DEBT_MDN    | Numeric  | $         |The median debt for students who have completed                       |   Financial Aid                 |
| WDRAW_DEBT_MDN   | Numeric  | $         |The median debt for sutdents who have not completed                   |Financial Aid |
| MD_EARN_WNE_P6   | Numeric  | $         |Median income six years after entry                                   |     earning   |
| MD_EARN_WNE_P10  | Numeric  | $         |Median income ten years after entry                                   |     earning  |
| RPY_3YR_RT_SUPP  | Numeric  | %         |Repayment rate 3 years after entering repayment                       |Repayment   |
| CDR3             | Numeric  | %         |Loan default rate 3 years afer entering repayment                     |Repayment        |
| PREDDEG_factor    | Factor  | N/A          |Predominant degree offered                                            |School       |
| Control_factor    | Factor  | N/A          |Control of institution                                                |School     |
| STABBR_factor  | Factor     | N/A       |State                                                                 |School     |
| APPL_SCH_PCT_GE2 | Numeric  | %         |Percentage of students submitting 2 or more than 2 applications       | Student        |
| UNITID | Numeric | N/A      | Unique ID for the institutions | unknown |
| INSTNM | Factor |  N/A      |name for the institutions       | unknown |

@@

Our objective in this stage is to conduct an exploration analysis on the CollegeScore dataset, investigating the individual variable's summary statistics and the relationship between these variabls, so as to better prepare for further analaysis on the factors affecting students' post college earnings with some advanced data mining techniques.

@@

summarize(data_clean,
          minPAR_ED_PCT_1STGEN=min(PAR_ED_PCT_1STGEN*100 ,na.rm=TRUE),
          maxPAR_ED_PCT_1STGEN=max(PAR_ED_PCT_1STGEN*100,na.rm=TRUE),
          meanPAR_ED_PCT_1STGEN=mean(PAR_ED_PCT_1STGEN*100,na.rm=TRUE),
          medianPAR_ED_PCT_1STGEN=median(PAR_ED_PCT_1STGEN*100,na.rm=TRUE),
          sdPAR_ED_PCT_1STGEN = sd(PAR_ED_PCT_1STGEN*100,na.rm=TRUE))

```
From the output, we can tell that the highest value existed in  La Belle Beauty Academy with the value of 95.25%. The lowest value is 6.87%, which existed in Wheaton College. From the mean value of 48.08%, we can tell that about for federally aided students in most colleges, around half (50.8%) of the students' parents did not complete the college. The standard deviation value of 12.2% indicated a relative high difference of this element among different colleges.

@@

  ggplot(data=data_clean , aes(PAR_ED_PCT_1STGEN)) + 
    geom_histogram(breaks=seq(0.1, 1, by =0.1), 
                   col="white", 
                   aes(fill=..count..),na.rm=TRUE)+
    labs(title="Histogram for PAR_ED_PCT_1STGEN") +
    labs(x="PAR_ED_PCT_1STGEN", y="Count") +
    theme(plot.title = element_text(hjust = 0.5))
```
From the histogram, we can tell the share of first generarion students is most frequently falling into the range between 0.5 and 0.6. The value with the value more than 0.8 are rare. The colleges with the share of first generarion students value less than 0.5 are much more than the ones with the value larger than 0.5. Thus the histogram is seriously left skewed.

@@

  #histogram for NPT_COM
  ggplot(data=data_clean , aes(NPT4_COMBINE)) + 
    geom_histogram(breaks=seq(0, 90000, by =2000), 
                   col="red", 
                   aes(fill=..count..),na.rm=TRUE)+
    labs(title="Histogram for NPT_COM") +
    labs(x="NPT_COM", y="Count") +
    theme(plot.title = element_text(hjust = 0.5))
```
From the histogram, we notice that the average net cost for IV institutions typically is 16000 -18000. The values between 16000 - 20000 are quite frequent. There were an unsually lagrge number of costs. The institutions with cost of more than 50000 were atypical. Due to the atypical large value, the histogram is abviously right skewed. Without those extreme values, the histogram would be reasonably symmetric.

@@

  #histogram for NPT_COM
  ggplot(data=data_clean , aes(DEP_STAT_PCT_IND)) + 
    geom_histogram(breaks=seq(0.1, 1, by =0.1), 
                   col="white", 
                   aes(fill=..count..),na.rm=TRUE)+
    labs(title="Histogram for DEP_STAT_PCT_IND") +
    labs(x="DEP_STAT_PCT_IND", y="Count") +
    theme(plot.title = element_text(hjust = 0.5))
```
From the histogram, we notice that the value between 0.7 to 0.8 are quite frequent. There was no obvious extrme value in the sample, thus the histogram is normally distributed.

@@

This elemnt, pulled from the IPEDS student Financial Aid Component, shows the share of undergraduate students who received Pell Grants in a given year, which is an important measure of the access an insiitution provides to low-income students.

We are using the code below to summarize this variable and extraced the instition with the highest and lowest value.

@@

  ggplot(data=data_clean , aes(PCTPELL)) + 
    geom_histogram(breaks=seq(0, 1, by =0.1), 
                   col="white", 
                   aes(fill=..count..),na.rm=TRUE)+
    labs(title="Histogram for PCTPELL") +
    labs(x="PCTPELL", y="Count") +
    theme(plot.title = element_text(hjust = 0.5))
```
From the histogram, we are able to conclude that in most colleges, 40% to 50% of the undergraduate students are able to receive the Pell Grant. With little skewness, the histogram is resonalbe symmetric.

@@

  ggplot(data=data_clean , aes(PCTFLOAN)) + 
    geom_histogram(breaks=seq(0, 1, by =0.1), 
                   col="white", 
                   aes(fill=..count..),na.rm=TRUE)+
    labs(title="Histogram for PCTFLOAN") +
    labs(x="PCTFLOAN", y="Count") +
    theme(plot.title = element_text(hjust = 0.5))
```
From the histogram, it's obvious that there are two peaks in the graph. The rates between 0- 10% and the rates between 60% to 80% are similarly frequent compated to the other ranges.

@@

This element is the median loan debt acculated at the institution by all student borrowers of federal lonans who have habe completed the education. We are using the code below to summarize this variable. 
```{R, message=FALSE, warinig = FALSE}

@@

 ggplot(data=data_clean , aes(GRAD_DEBT_MDN)) + 
    geom_histogram(breaks=seq(0, 50000, by =2000), 
                   col="red", 
                   aes(fill=..count..),na.rm=TRUE)+
    labs(title="Histogram for GRAD_DEBT_MDN") +
    labs(x="GRAD_DEBT_MDN", y="Count") +
    theme(plot.title = element_text(hjust = 0.5))

@@

###5.1.5 Earning<a id="5.1.5"></a> 
####5.1.5.1 Median family income in real 2015 dollars<a id="5.1.5.1"></a>
This elemnt is recording the median family income of students, one of the undergraduate student demographics for earnings cohorts.

We are using the code below to summarize this variable.

@@

  ggplot(data=data_clean , aes(MD_FAMINC)) + 
    geom_histogram(breaks=seq(0, 156000, by =3000), 
                   col="red", 
                   aes(fill=..count..),na.rm=TRUE)+
    labs(title="Histogram for MD_FAMINC") +
    labs(x="MD_FAMINC", y="Count") +
    theme(plot.title = element_text(hjust = 0.5))
```

@@

##6.3. The relationship between Percentage of Pell Grant and college type(CONTROL)<a id="6.3"></a>

To compare the percentages of Pell Grant in different control type, we grouped the Percentage of Pell Grant data into the college type level to take a further analysis. We would like to investigate the percentage of Pell Grant based on the  college type.

@@

ggplot(data_clean, aes(x=PCTPELL, fill=Control_factor)) + geom_histogram(alpha=0.9, position="identity") +
  labs(title="Overlaying histogram between PCTPELL and Control ") +
    theme(plot.title = element_text(hjust = 0.5))

@@

From the result, the Private institutions have a much larger median value than the public institutions. With a much higher costs burden, it makes sence that the low-income sutdents in private institutions are much more likely to seek financial aid than the public institution students. 
From the stacked histogram, we can tell that for all the three college type institutions, the most frequent ranges appear between 0.6 and 0.8, which is consistent with the findings we observe in the individual variable summary section.

@@

>>>>

The data is collected by United States Department of Education, which corresponds to all undergraduate degree-granting institutions of higher education from 1996 through 2016. The dataset is pretty rich and covers a wide range of topics. It has 1743 associated columns. Each corresponds to one named data variable. All data variables are divided into 10 categories ranging from Root category to  Earnings category. The data type contains integer, string and float. Data elements with an IPEDS value in the Source field are derived directly from the Integrated Postsecondary Education Data System (IPEDS). For further detailed description of the dataset, you can find it in the CollegeScorecardDataDictionary.xlsx excel file.

@@

@@

for (i in 1:length(data_new$NPT4_PRIV)) {
   if (is.na(data_new$NPT4_PRIV[i])) {
       data_new$NPT4_PRIV[i] = data_new$NPT4_PUB[i]
   }
}

mean(is.na(data_new$NPT4_PRIV))
data_new = data_new[,-10] 
colnames(data_new)[10] <- "NPT4_COMBINE"

@@

In order to choose the variables that have enough valid data related to the topic, we filtered and considered variables with percent missing value less than 25%. The new data set was named as data_clean for further analysis. 
```{r, message=FALSE,warning=FALSE}

@@

For numeric variabls, we calculate the basic summary statistics such as median, min, max, mode and quantile. We also use visualizations such as hisgroms and boxplots to understand the distribution of the values. For categorical variables, we use frequency table to understand the frequcies of each categories. Box chart is used as a visualization tool.

The table below lists the variables that will be used for the further analysis: 13 of them are continuous variables while 5 of them are categorical variables or Unique ID. After initial data cleaning process, the new data set now contain 18 variables and 7793 observations.

| Variable Name    | Data Type |  Unit  |  Description                                                          |    Category in Documentation       |
| --------------   | ----------|--------|-----------------------------------------------------------------------| --------------------------------|
| NPT4_COMBINE    | Numeric   | $         |average net price for TITLE IV institutions of both public and private.|  Cost                       |
| DEP_STAT_PCT_IND | Numeric  | %         |Percentage of students who are financially independent                |     Cost                        |
| PAR_ED_PCT_1STGEN| Numeric  | %         |Percentage first-generation students                                  |      Sudent Body                |
| MD_FAMINC        | Numeric  | $         |Median family income in real 2015 dollars                             |       Earning                   |
| PCTPELL          | Numeric  | %         |Percentage of undergraduates who receive a Pell Grant                 |        Financial Aid            |
| PCTFLOAN         | Numeric  | %         |Percent of all federal undergraduate students receiving a federal student loan      |  Financial Aid    |
| GRAD_DEBT_MDN    | Numeric  | $         |The median debt for students who have completed                       |   Financial Aid                 |
| WDRAW_DEBT_MDN   | Numeric  | $         |The median debt for sutdents who have not completed                   |Financial Aid |
| MD_EARN_WNE_P6   | Numeric  | $         |Median income six years after entry                                   |     earning   |
| MD_EARN_WNE_P10  | Numeric  | $         |Median income ten years after entry                                   |     earning  |
| RPY_3YR_RT_SUPP  | Numeric  | %         |Repayment rate 3 years after entering repayment                       |Repayment   |
| CDR3             | Numeric  | %         |Loan default rate 3 years afer entering repayment                     |Repayment        |
| PREDDEG_factor    | Factor  | N/A          |Predominant degree offered                                            |School       |
| Control_factor    | Factor  | N/A          |Control of institution                                                |School     |
| STABBR_factor  | Factor     | N/A       |State                                                                 |School     |
| APPL_SCH_PCT_GE2 | Numeric  | %         |Percentage of students submitting 2 or more than 2 applications       | Student        |
| UNITID | Numeric | N/A      | Unique ID for the institutions | Root |
| INSTNM | String |  N/A      |name for the institutions       | Institution |

@@

Our objective in this stage is to conduct an exploration analysis on the CollegeScore dataset, investigating the individual variable's summary statistics and the relationship between these variabls, so as to better prepare for further analaysis on the factors affecting students' post college earnings with some advanced data mining techniques. The summary statistics, dplyr function, ggplot functions as well as some visualization techniques will be mainly used in this project.

@@

#summarize the variable of PAR_ED_PCT_1STGEN
summarize(data_clean,
          minPAR_ED_PCT_1STGEN=min(PAR_ED_PCT_1STGEN*100 ,na.rm=TRUE),
          maxPAR_ED_PCT_1STGEN=max(PAR_ED_PCT_1STGEN*100,na.rm=TRUE),
          meanPAR_ED_PCT_1STGEN=mean(PAR_ED_PCT_1STGEN*100,na.rm=TRUE),
          medianPAR_ED_PCT_1STGEN=median(PAR_ED_PCT_1STGEN*100,na.rm=TRUE),
          sdPAR_ED_PCT_1STGEN = sd(PAR_ED_PCT_1STGEN*100,na.rm=TRUE))

```
From the output, we can tell that the highest value existed in  La Belle Beauty Academy with the value of 95.25%. The lowest value is 6.87%, which existed in Wheaton College. From the mean value of 48.08%, we can tell that for federally aided students in most colleges, around half (50.8%) of the students' parents did not complete the college. The standard deviation value of 12.2% indicated a relative high difference of this element among different colleges.

@@

  ggplot(data=data_clean , aes(PAR_ED_PCT_1STGEN)) + 
    geom_histogram(breaks=seq(0.1, 1, by =0.1), 
                   col="white", 
                   aes(fill=..count..),na.rm=TRUE)+
    labs(title="Histogram for Percentage first-generation students") +
    labs(x="PAR_ED_PCT_1STGEN", y="Count") +
    theme(plot.title = element_text(hjust = 0.5))
```
From the histogram, we can tell the share of first generarion students is most frequently falling into the range between 0.5 and 0.6. The value with the value more than 0.8 are rare. The colleges with the share of first generarion students value less than 0.5 are much more than the ones with the value larger than 0.5. Thus the histogram is seriously left skewed.

@@

  #histogram for NPT_COM
  ggplot(data=data_clean , aes(NPT4_COMBINE)) + 
    geom_histogram(breaks=seq(0, 90000, by =2000), 
                   col="red", 
                   aes(fill=..count..),na.rm=TRUE)+
    labs(title="Histogram for Average Net Price for Title IV institutions") +
    labs(x="Average Net Price", y="Count") +
    theme(plot.title = element_text(hjust = 0.5))
```
From the histogram, we notice that the average net cost for IV institutions typically is $16000 - $18000. The values between $16000 - $20000 are quite frequent. There were an unsually lagrge number of costs. The institutions with cost of more than 50000 were atypical. Due to the atypical large value, the histogram is abviously right skewed. Without those extreme values, the histogram would be reasonably symmetric.

@@

  #histogram for NPT_COM
  ggplot(data=data_clean , aes(DEP_STAT_PCT_IND)) + 
    geom_histogram(breaks=seq(0.1, 1, by =0.1), 
                   col="white", 
                   aes(fill=..count..),na.rm=TRUE)+
    labs(title="Histogram for Percentage of Students Who are Financially Independent") +
    labs(x="DEP_STAT_PCT_IND", y="Count") +
    theme(plot.title = element_text(hjust = 0.5))
```
From the histogram, we notice that the value between 0.7 to 0.8 are quite frequent. There was no obvious extrme value in the sample, thus the histogram is normally distributed.

@@

This elemnt, pulled from the IPEDS student Financial Aid Component, shows the share of undergraduate students who received Pell Grants in a given year, which is an important measure of the access an insiitution provides to low-income students. We are using the code below to summarize this variable and extraced the instition with the highest and lowest value.

@@

  ggplot(data=data_clean , aes(PCTPELL)) + 
    geom_histogram(breaks=seq(0, 1, by =0.1), 
                   col="white", 
                   aes(fill=..count..),na.rm=TRUE)+
    labs(title="Histogram for Percentage of Undergraduates who Receive a Pell Grant") +
    labs(x="PCTPELL", y="Count") +
    theme(plot.title = element_text(hjust = 0.5))
```
From the histogram, we are able to conclude that in most colleges, 40% to 50% of the undergraduate students are able to receive the Pell Grant. With little skewness, the histogram is resonalbe symmetric.

@@

  ggplot(data=data_clean , aes(PCTFLOAN)) + 
    geom_histogram(breaks=seq(0, 1, by =0.1), 
                   col="white", 
                   aes(fill=..count..),na.rm=TRUE)+
    labs(title="Histogram for Percent of All Federal Undergraduate Students Receiving a Federal Student Loan") +
    labs(x="PCTFLOAN", y="Count") +
    theme(plot.title = element_text(hjust = 0.5))
```
From the histogram, it's obvious that there are two peaks in the graph. The rates between 0- 10% and the rates between 60% to 80% are similarly frequent compated to the other ranges.

@@

This element is the median loan debt acculated at the institution by all student borrowers of federal lonans who have habe completed the education. We are using the code below to summarize this variable??? extracting the largest and lowest value for this variable.
```{R, message=FALSE, warinig = FALSE}

@@

 ggplot(data=data_clean , aes(GRAD_DEBT_MDN)) + 
    geom_histogram(breaks=seq(0, 50000, by =2000), 
                   col="red", 
                   aes(fill=..count..),na.rm=TRUE)+
    labs(title="Histogram for median debt for students who have completed") +
    labs(x="GRAD_DEBT_MDN", y="Count") +
    theme(plot.title = element_text(hjust = 0.5))

@@

###5.1.5 Earning<a id="5.1.5"></a> 
####5.1.5.1 Median family income in real 2015 dollars<a id="5.1.5.1"></a>
This elemnt is recording the median family income of students, one of the undergraduate student demographics for earnings cohorts. We are using the code below to summarize this variable.

@@

  ggplot(data=data_clean , aes(MD_FAMINC)) + 
    geom_histogram(breaks=seq(0, 156000, by =3000), 
                   col="red", 
                   aes(fill=..count..),na.rm=TRUE)+
    labs(title="Histogram for Median family income in real 2015 dollars") +
    labs(x="MD_FAMINC", y="Count") +
    theme(plot.title = element_text(hjust = 0.5))
```

@@

##6.3. The relationship between Percentage of Pell Grant and College Type(CONTROL)<a id="6.3"></a>

To compare the percentages of Pell Grant in different control type, we grouped the data into the college type level to take a further analysis.

@@

Below is the stacked histogram between these two variables. 
```{r}

ggplot(data_clean, aes(x=PCTPELL, fill=Control_factor)) + geom_histogram(alpha=0.9, position="identity") +
  labs(title="Stacked histogram between PCTPELL and Control ") +
    theme(plot.title = element_text(hjust = 0.5))

@@

From the result, the Private institutions have a much larger median value than the public institutions. With a much higher costs burden, it makes sense that the low-income sutdents in private institutions are much more likely to seek financial aid than the public institution students.

From the stacked histogram, we can tell that for all the three college type institutions, the most frequent ranges appear between 0.6 and 0.8, which is consistent with the findings we observe in the previous individual variable summary section.

@@

Through the process of summarizing these variables and analyzing the relationship between these variables, we found that for federally aided students in most colleges, around half (50.8%) of the students' parents did not complete the college.The average net price costs for most institutuins are falling into the range between $16000 - $20000. Among the 7703 institution samples, more than 50% of the students are able to receive Pell Grant or Federal Loans as means of finanical support??? and on average, 60% of students are financial independent. In most colleges, the students are able to reach the family income between $12000 and $18000.

Our findings also suggest that the Title IV institions located in DC, VT, MA, NH, RI have a relative high amount of average net costs than those in other states. The students in private non-profit institutions and public institutions are less likely to seek Pell Grant as financial assistance than the students in the private for-profit colleges.In most private for-profit colleges, the students are more likely to be financial independet.

It is also indicated that there exists a negative correlation between the percentage of students who receive Pell Grant and the students median family income after they graduate: for institutions with low percentage of students who receive the Pell Grant, the financial aids like Pell Grant may not be as necessary as they are in the instituions with high value of PCTPELL, thus these collges students will tend to have a high median family income after they graduate.

As a result of all the statistical analysis, we conclude that XXXXXXXXXXXX. Attributes such as                will affec the students' post graudate earnings.
